On the Bit Complexity of Finding Points in Connected Components of a Smooth Real Hypersurface

### 2020-02-07 main.tex
Minor intro updates

  Our starting point is an algorithm by Safey El Din and Schost ({\em
    Polar varieties and computation of one point in each connected
    component of a smooth real algebraic set}, ISSAC'03). This
  algorithm uses random changes of variables, that are proved to
  generically ensure certain desirable geometric properties. The
  complexity of the algorithm was given in an algebraic complexity
  model; the analysis of the bit complexity and the error probability
  were left for future work.

@@

The number of connected components of $V(f)$ admits the lower bound
$d^{\Omega(n)}$, so up to polynomial factors, this result is
optimal. However, due to the generality of the algorithm, the constant
hidden in the exponent $O(n)$ in its runtime turns out to be rather
large: the algorithm relies on infinitesimal deformations, that affect
runtime non-trivially.

@@

Any polynomial of degree $d$ in $n$ variables can be computed by a
straight-line program of length $O\big ({{d+n} \choose n}\big)$:
evaluate all monomials of degree up to $d$ in $n$ variables, multiply
them by their respective coefficients and sum the result. However,
some inputs may be given by shorter straight-line program, and the
algorithm is able to benefit from it.

@@

  As part of the randomization, the algorithm chooses a set 
  \[
  S := \{1,2,\hdots,2\epsilon^{-1}3n^2(2d)^{2n}\}
  \]
  from which to sample from.

@@

The runtime for cases without a strait-line program in the input are as follows. In these cases a polynomial $f$ of degree $d$ is given in the input, and a strait-line program is built for it by computing all the terms. In this case, $L \geq d^n$ and $b \geq \htt(f) = s,$ so that the number of bit operations becomes
\begin{align*}
&O^{\sim}(nk(d^ns+d^{2n}(s + \log 1/\epsilon+d)(d^n+d)\log(s + \log 1/\epsilon)))\\
      &= 
O^{\sim}(nk(d^ns+(d^{3n}+d^{2n+1})(s + \log 1/\epsilon+d)\log(s + \log 1/\epsilon)))\\
     &= 
O^{\sim}(nk(d^{3n}(s + \log 1/\epsilon+d)\log(s + \log 1/\epsilon)))\\
     &= 
O^{\sim}(nk(d^{3n+1}(s + \log 1/\epsilon)\log(s + \log 1/\epsilon))),
\end{align*}
and the degree and height of the output does not change.

@@

Our main contribution is to analyse precisely what conditions on our
change of coordinates $\mA$ guarantee success. This is done by
revisiting the key ingredients in the proofs given
in~\cite{BaGiHeMb97} and~\cite{EMP} and giving quantitative versions
of these results, bounding the degree of the hypersurfaces we have to
avoid.  To solve the equations~\eqref{eq:syst}, we use the algorithm
in~\cite{SH}, for which a complete bit complexity analysis is
available.

This work should be seen as a first step toward the analysis of
further randomized algorithms in real algebraic geometry. An immediate
follow-up question would be to handle the case of algebraic sets
defined by {\em regular sequences}: the algorithm in~\cite{EMP} still
applies, but the modifications needed would take us to far for the
current format. Further yet, randomized algorithms for deciding {\em
  connectivity queries} on smooth, compact algebraic sets have been
developed in a series of papers
\cite{SchostMohabBabySteps2011,SchostMohabBabySteps2014}, and could be
revisited using the techniques introduced here.

@@

  With probability at least $1-\epsilon$, the output is correct. Otherwise, the algorithm either returns a proper subset of the points or FAIL. In any case, the algorithm uses
  \[
  O^{\sim}(nk(Lb+d^{2n}(s + \log 1/\epsilon+d)(L+d)\log(s + \log 1/\epsilon)))
  \]
  bit operations, with $k \geq 1+\lg n + \lg(1/\epsilon)$. The polynomials in
  the output have degree at most $d^n,$ and height 
  \[
  O^{\sim}((s+\lg 1/\epsilon + d)d^n)
  =O^{\sim}((s+\lg 1/\epsilon)d^{n+1}).
  \]
%
\subsection{The Algorithm}
%
For a matrix $A \in \GL(n,\Q),$ let 
\[
Y_i = X_i^{A} = [b_i] \cdot \bbm X_1\\ \vdots \\ X_n \ebm, 
\]
where $b_i$ is the i-th row of $A^{-1}$.
%
\begin{algorithm}[hbt!]
\KwIn{a strait line program $\Gamma$ that computes $f \in \Q[X_1,\hdots,X_n],$ and $d=\deg(f)$}
\KwOut{$n$ zero-dimensional parameterizations, the union of which includes at least one point in each connected component of $V(f) \cap \R^n$}
\nl Take  $S := \{1,2,\hdots,2\epsilon^{-1}3n^2(2d)^{3n}\}$
 and randomly choose $A \in S^{n^2}$, and  $\sigma_{n-1}=(\sigma_1,\hdots,\sigma_{n-1}) \in S^{n-1}$\;
\nl Compute $f$ and apply the change of variables to obtain $f(A x)=f^A(x)$\;
\nl Use \cite{partialDerComputation} to compute the partial derivatives $\frac{\pa f^A}{\pa X_{i+1}},\hdots,\frac{\pa f^A}{\pa X_n}$\;
    \caption{{Main Algorithm} \label{}}
\nl \For{$i\gets1$ \KwTo $n$}{
\nl Run \cite[Algorithm 2]{SH} $k \geq 1+\lg n + \lg(1/\epsilon)$ times with input (state this in terms of straight line programs?)
$\{Y_1-\sigma_1,\hdots,Y_{i-1}-\sigma_{i-1},f^A,\frac{\pa f^A}{\pa X_{i+1}},\hdots,\frac{\pa f^A}{\pa X_n} \} \textrm{ and } d$\;
\nl Let $\mathscr{Q}_i$ be the highest cardinality zero-dimensional parameterization returned in step 5 \;
} 
\nl \Return $[\mathscr{Q}_1,\hdots,\mathscr{Q}_n]$.
\end{algorithm}

@@

Let $\mathscr{E}$ be the event that, the parameterizations $[\mathscr{Q}_1,\hdots,\mathscr{Q}_n]$ returned in step 7 of Algorithm 1 contain all points and are correct. Then, 
\begin{align*}
\pr[\textrm{success}] = \pr[A \not\in \Delta] \times \pr[\mathscr{E}~|~A \not\in \Delta].  
\end{align*}
By construction of $S$,
\[
\pr[A \not\in \Delta] = 1-\pr[A \in \Delta] \geq
1-\epsilon/2.
\]
Now set $k = \lg(2n/\epsilon)$ so that 
\[
(1-2^{-k})^n = (1 - \epsilon/(2n))^n \geq 1 - \epsilon/2,
\]
by Bernoulli's inequality. Therefore, 
\begin{align*}
\pr[\textrm{success}] &\geq  (1-\epsilon/2)\pr[\mathscr{E}~|~A \not\in \Delta]\\
&\geq (1-\epsilon/2)(1-(11/32)^k)^n\\
& \geq (1-\epsilon/2)(1-2^{-k})^n \\
& \geq (1-\epsilon/2)(1-\epsilon/2)\\
& = 1 - \epsilon + \epsilon^2/4 \geq 1- \epsilon.
\end{align*}
%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}
%%
\end{document}
\endinput

>>>>

  Our starting point is an algorithm by Safey El Din and Schost ({\em
    Polar varieties and computation of one point in each connected
    component of a smooth real algebraic set}, ISSAC'03). This
  algorithm uses random changes of variables, that are proved to
  generically ensure certain desirable geometric properties. The
  cost of the algorithm was given in an algebraic complexity
  model; the analysis of the bit complexity and the error probability
  were left for future work.

@@

The number of connected components of $V(f)$ admits the lower bound
$d^{\Omega(n)}$, so up to polynomial factors this result is
optimal. However, due to the generality of the algorithm, the constant
hidden in the exponent $O(n)$ in its runtime turns out to be rather
large: the algorithm relies on infinitesimal deformations, that affect
runtime non-trivially.

@@

Any polynomial of degree $d$ in $n$ variables can be computed by a
straight-line program of length $O\big ({{d+n} \choose n}\big)$:
evaluate all monomials of degree up to $d$ in $n$ variables, multiply
them by their respective coefficients and sum the result. However,
some inputs may be given by shorter straight-line program, and the
algorithm is able to benefit from this.

@@

  As part of the randomization, the algorithm chooses a set 
  \[
  S := \{1,2,\hdots,2\epsilon^{-1}3n^2(2d)^{2n}\}
  \]
  from which to sample.

@@

The runtime for cases without a straight-line program in the input are as follows. In these cases a polynomial $f$ of degree $d$ is given in the input, and a straight-line program is built for it by computing all the terms. In this case, $L \geq d^n$ and $b \geq \htt(f) = s,$ so that the number of bit operations becomes
\begin{align*}
&O^{\sim}(nk(d^ns+d^{2n}(s + \log 1/\epsilon+d)(d^n+d)\log(s + \log 1/\epsilon)))\\
      &= 
O^{\sim}(nk(d^ns+(d^{3n}+d^{2n+1})(s + \log 1/\epsilon+d)\log(s + \log 1/\epsilon)))\\
     &= 
O^{\sim}(nk(d^{3n}(s + \log 1/\epsilon+d)\log(s + \log 1/\epsilon)))\\
     &= 
O^{\sim}(nk(d^{3n+1}(s + \log 1/\epsilon)\log(s + \log 1/\epsilon))),
\end{align*}
and the degree and height of the output does not change.

@@

Our main contribution is to analyse precisely what conditions on our
change of coordinates $\mA$ guarantee success. This is done by
revisiting the key ingredients in the proofs given
in~\cite{BaGiHeMb97} and~\cite{EMP}, and giving quantitative versions
of these results, bounding the degree of the hypersurfaces we have to
avoid.  To solve the equations~\eqref{eq:syst}, we use the algorithm
in~\cite{SH}, for which a complete bit complexity analysis is
available.

This work should be seen as a first step toward the analysis of
further randomized algorithms in real algebraic geometry. An immediate
follow-up question would be to handle the case of algebraic sets
defined by {\em regular sequences}: the algorithm in~\cite{EMP} still
applies, but the modifications needed are beyond the scope of this
publication. Further still, randomized algorithms for deciding {\em
  connectivity queries} on smooth, compact algebraic sets have been
developed in a series of papers
\cite{SchostMohabBabySteps2011,SchostMohabBabySteps2014}, and could be
revisited using the techniques introduced here.

@@

  With probability at least $1-\epsilon$, the output is correct. Otherwise, the algorithm either returns a proper subset of the points or FAIL. In any case, the algorithm uses
  \[
  O^{\sim}(nk(Lb+d^{2n}(s + \log 1/\epsilon+d)(L+d)\log(s + \log 1/\epsilon)))
  \]
  bit operations, with $k \geq 1+\lg n + \lg(1/\epsilon)$. The polynomials in
  the output have degree at most $d^n,$ and height 
  \[
  O^{\sim}((s+\lg 1/\epsilon + d)d^n)
  =O^{\sim}((s+\lg 1/\epsilon)d^{n+1}).
  \]
%
\subsection{The Algorithm}
%
For a matrix $A \in \GL(n,\Q),$ let 
\[
Y_i = X_i^{A} = [b_i] \cdot \bbm X_1\\ \vdots \\ X_n \ebm, 
\]
where $b_i$ is the i-th row of $A^{-1}$.
%
\begin{algorithm}[hbt!]
\KwIn{a straight line program $\Gamma$ that computes $f \in \Q[X_1,\hdots,X_n],$ and $d=\deg(f)$}
\KwOut{$n$ zero-dimensional parameterizations, the union of which includes at least one point in each connected component of $V(f) \cap \R^n$}
\nl Take  $S := \{1,2,\hdots,2\epsilon^{-1}3n^2(2d)^{3n}\}$
 and randomly choose $A \in S^{n^2}$, and  $\sigma_{n-1}=(\sigma_1,\hdots,\sigma_{n-1}) \in S^{n-1}$\;
\nl Compute $f$ and apply the change of variables to obtain $f(A x)=f^A(x)$\;
\nl Use \cite{partialDerComputation} to compute the partial derivatives $\frac{\pa f^A}{\pa X_{i+1}},\hdots,\frac{\pa f^A}{\pa X_n}$\;
    \caption{{Main Algorithm} \label{}}
\nl \For{$i\gets1$ \KwTo $n$}{
\nl Run \cite[Algorithm 2]{SH} $k \geq 1+\lg n + \lg(1/\epsilon)$ times with input (state this in terms of straight line programs?)
$\{Y_1-\sigma_1,\hdots,Y_{i-1}-\sigma_{i-1},f^A,\frac{\pa f^A}{\pa X_{i+1}},\hdots,\frac{\pa f^A}{\pa X_n} \} \textrm{ and } d$\;
\nl Let $\mathscr{Q}_i$ be the highest cardinality zero-dimensional parameterization returned in step 5 \;
} 
\nl \Return $[\mathscr{Q}_1,\hdots,\mathscr{Q}_n]$.
\end{algorithm}

@@

Let $\mathscr{E}$ be the event that, the parameterizations $[\mathscr{Q}_1,\hdots,\mathscr{Q}_n]$ returned in step 7 of Algorithm 1 contain all points and are correct. Then, 
\begin{align*}
\pr[\textrm{success}] = \pr[A \not\in \Delta] \times \pr[\mathscr{E}~|~A \not\in \Delta].  
\end{align*}
By construction of $S$,
\[
\pr[A \not\in \Delta] = 1-\pr[A \in \Delta] \geq
1-\epsilon/2.
\]
Now set $k = \lg(2n/\epsilon)$ so that 
\[
(1-2^{-k})^n = (1 - \epsilon/(2n))^n \geq 1 - \epsilon/2,
\]
by Bernoulli's inequality. Therefore, 
\begin{align*}
\pr[\textrm{success}] &\geq  (1-\epsilon/2)\pr[\mathscr{E}~|~A \not\in \Delta]\\
&\geq (1-\epsilon/2)(1-(11/32)^k)^n\\
& \geq (1-\epsilon/2)(1-2^{-k})^n \\
& \geq (1-\epsilon/2)(1-\epsilon/2)\\
& = 1 - \epsilon + \epsilon^2/4 \geq 1- \epsilon.
\end{align*}
%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}
%%
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
